\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{parskip}

\title{Chess Thermodynamics:\\A Gibbs Free Energy Framework for Positional Evaluation}
\author{}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We develop a chess evaluation framework grounded in statistical mechanics.
Each chess position is treated as a thermodynamic state characterized by
internal energy~$U$, temperature~$T$, and entropy~$S$.
The evaluation decomposes into a \emph{free energy} component
$F = U - TS$ that captures the kinetic--entropic balance of each side,
and a \emph{potential energy} component $PV$ fitted from master games
that encodes structural and tactical interactions.
The engine evaluation is $\mathcal{E} = (\Delta F - PV)/\bar{T}$,
expressed in piece-equivalent units.
At thermodynamic equilibrium---well-played positions by strong players---the
Gibbs relation $G = U + PV - TS \approx 0$ holds, providing the fitting
constraint for the potential energy weights.
\end{abstract}

\section{Motivation}

Traditional chess engines evaluate positions by assigning fixed point
values to material and adding hand-tuned bonuses for positional features
(king safety, pawn structure, piece activity, etc.). This approach is
effective but theoretically ad hoc: there is no principled relationship
between the material term and the positional corrections.

Statistical mechanics offers a natural language for this problem. A chess
position is a \emph{configuration} of interacting particles (pieces) on a
lattice (the board). Each piece has degrees of freedom (squares it can
reach), and the collective behavior of the position---how well-coordinated
the pieces are, how efficiently mobility is distributed, how much the
opposing forces overlap---maps onto thermodynamic quantities.

The key insight is that well-played games sit near \emph{thermodynamic
equilibrium}: strong players maintain a balance between energy, entropy,
and structural tension. Deviations from equilibrium correspond to
advantages that one side can exploit.

\section{Thermodynamic Variables}

\subsection{Internal Energy: $U$}

The internal energy of a side is its \textbf{total x-ray mobility}---the
number of squares each piece could reach if no other pieces blocked its
path, summed over all pieces.

\begin{equation}
U = \sum_{i \in \text{pieces}} f_i
\end{equation}

where $f_i$ is the x-ray mobility of piece~$i$ (the number of squares it
attacks or moves to on an empty board from its current square, following
its movement rules along full rays).

X-ray mobility is a function of piece type and square alone, independent of
the current board configuration. This makes $U$ a measure of the
\emph{intrinsic energy potential} of the pieces---what they could
accomplish if unobstructed. A queen on d4 contributes $\sim$27 squares; a
pawn on e2 contributes $\sim$4. Material advantage is directly reflected:
removing a piece always reduces $U$ for that side.

\subsection{Temperature: $T$}

Temperature measures the \textbf{concentration of available movement} among
the pieces. It is defined as:

\begin{equation}
T = \frac{M}{N_{\text{eff}}}, \qquad N_{\text{eff}} = e^{H}
\end{equation}

where $M = \sum_i m_i$ is the total number of \emph{legal} moves
(accounting for blocking, pins, etc.), $m_i$ is the legal move count of
piece~$i$, and $H$ is the Shannon entropy of the per-piece legal move
distribution:

\begin{equation}
H = -\sum_{i} \frac{m_i}{M} \ln \frac{m_i}{M}
\end{equation}

$N_{\text{eff}} = e^H$ is the effective number of pieces contributing to
the legal move pool. Temperature captures the \emph{game phase}:

\begin{itemize}[nosep]
\item \textbf{Cold positions} (middlegame): Legal moves are spread across
  many pieces. $N_{\text{eff}}$ is large, so $T = M/N_{\text{eff}}$ is
  small. The position has high ``heat capacity''---many degrees of freedom
  absorb the available energy.
\item \textbf{Hot positions} (endgame): Legal moves are concentrated in a
  few active pieces. $N_{\text{eff}}$ is small, so $T$ is large. Each
  remaining piece carries a disproportionate share of the available
  mobility.
\end{itemize}

This matches chess intuition: in endgames, each piece matters more
(higher temperature), and material advantages are amplified.

\subsection{Entropy: $S$}

Entropy measures \textbf{x-ray redundancy}---the degree to which the x-ray
coverage of friendly pieces overlaps.

\begin{equation}
S = 1 - \frac{|\text{union of x-ray squares}|}{\sum_i f_i}
\end{equation}

If every piece's x-ray squares were disjoint, the union would equal the
sum and $S = 0$ (no redundancy, no entropy). If many pieces cover the same
squares, the union is much smaller than the sum and $S$ approaches~1 (high
redundancy, high entropy).

Only friendly pieces contribute to each side's entropy. Opponent overlap is
a distinct physical effect captured by the potential energy.

The analogy to statistical mechanics is precise: $S$ plays the role of
\emph{configurational entropy} arising from excluded-volume interactions.
Just as gas molecules in a container have reduced phase space when their
excluded volumes overlap, chess pieces with overlapping x-ray coverage have
reduced ``effective'' degrees of freedom. This is a geometric interaction
(phase-space overlap), not an energetic one---which is why it appears in
$S$ rather than in the potential energy $PV$.

\subsection{Free Energy: $F = U - TS$}

The Helmholtz free energy per side is:

\begin{equation}
F = U - TS
\end{equation}

This represents the \emph{usable} portion of the internal energy after
subtracting the mobility that is ``lost to disorder''---dispersed across
redundant coverage. The free energy difference between white and black is:

\begin{equation}
\Delta F = F_w - F_b = (U_w - T_w S_w) - (U_b - T_b S_b)
\end{equation}

$\Delta F > 0$ means white has more usable energy; $\Delta F < 0$ favors
black.

\section{Potential Energy: $PV$}

The free energy $\Delta F$ captures the kinetic--entropic balance but does
not account for \emph{structural interactions} between pieces: how well
they coordinate attacks, control key squares, pressure the opposing king,
or support pawn advances. These are the energetic (as opposed to entropic)
piece-piece interactions, analogous to intermolecular forces in a real gas.

We model the potential energy as a function of 23 structural features
$\phi_k$ computed from the position, with both linear and quadratic
(interaction) terms:

\begin{equation}
PV = \sum_{k=1}^{23} w_k \, \phi_k + \sum_{k \le l} w_{kl} \, \phi_k \, \phi_l
\end{equation}

The structural features fall into five categories:

\subsubsection*{Field Multipoles (10 features)}

Each piece acts as a source with charge equal to its x-ray mobility~$f_i$.
These sources are propagated through a Green's function
$G(x, s) = 1/(1 + d_{\text{Chebyshev}}(x,s))$ to produce smooth field
densities $\rho_w(x)$ and $\rho_b(x)$ over the board.

The center of the combined field is computed, along with the inertia tensor
and its eigenvectors (anchored so the principal axis points toward rank~8).
In this eigenbasis, we extract per-side \textbf{dipole moments} ($d_1$,
$d_2$) and \textbf{quadrupole moments} ($q_{11}$, $q_{22}$, $q_{12}$),
yielding 10 features total.

These capture the large-scale spatial distribution of each side's
force---whether pieces project forward, concentrate on one flank, or
spread symmetrically.

\subsubsection*{King--Field Coupling (4 features)}

The propagated enemy field evaluated at each king's position measures
\textbf{attack pressure}: how strongly the opponent's pieces project force
toward the king. The friendly field at the king measures
\textbf{defensive shielding}. This yields four features: attack on white
king, defense of white king, attack on black king, defense of black king.

\subsubsection*{Meta-Entropy (2 features)}

The Shannon entropy of the per-piece x-ray source distribution:

\begin{equation}
\eta = -\sum_i \frac{f_i}{\sum_j f_j} \ln \frac{f_i}{\sum_j f_j}
\end{equation}

This measures how evenly the x-ray ``energy'' is distributed across
pieces. A side with one dominant piece (e.g., a lone queen) has low
meta-entropy; a side with many roughly equal pieces has high meta-entropy.
One feature per side.

\subsubsection*{Pawn Structure (6 features)}

Per-side counts of pawns, total pawn advancement (sum of ranks advanced
from starting position), and pawn file spread (variance of file positions).
These capture pawn chain structure, passed pawn potential, and spatial
coverage.

\subsubsection*{Spatial Tension (1 feature)}

The overlap integral of the opposing propagated fields:

\begin{equation}
\tau = \sum_{x} \rho_w(x) \cdot \rho_b(x)
\end{equation}

High tension indicates that the two sides' spheres of influence strongly
overlap---a hallmark of tactical positions with imminent contact.

\section{The Equation of State}

The total Gibbs free energy of the position is:

\begin{equation}
G = \Delta F - PV
\end{equation}

The engine evaluation, normalized to piece-equivalent units, is:

\begin{equation}
\boxed{\mathcal{E} = \frac{\Delta F - PV}{\bar{T}}}
\end{equation}

where $\bar{T} = (T_w + T_b)/2$ is the average temperature. The division
by $\bar{T}$ converts from energy units (squares) to piece-equivalent
units, ensuring that evaluations are comparable across game phases: a
two-piece advantage means the same thing whether the position is a
middlegame or an endgame.

The evaluation bar uses a logistic sigmoid $P = 1/(1 + e^{-\mathcal{E}})$,
mapping directly from piece advantage to an estimated win probability
without arbitrary scaling constants.

\section{Fitting the Potential Energy}

At thermodynamic equilibrium---positions from well-played games between
strong players---the Gibbs free energy vanishes:

\begin{equation}
G \approx 0 \quad \Longrightarrow \quad PV \approx \Delta F
\end{equation}

This provides the fitting constraint. From a database of master-level
games, we extract positions (skipping the opening), compute $\Delta F$ and
the 23 structural features for each, and fit the potential energy weights
via ridge regression:

\begin{equation}
\min_{\mathbf{w}} \sum_{\text{positions}} \left( \Delta F - PV(\mathbf{w}) \right)^2 + \alpha \|\mathbf{w}\|^2
\end{equation}

The linear and quadratic weights ($23 + 276 = 299$ parameters total) are
fitted jointly. Feature standardization (zero mean, unit variance) is
applied before regression to handle the wide dynamic range of the
structural features (spatial tension $\sim 10^4$ vs.\ pawn counts
$\sim 10^0$).

After fitting, $PV$ approximates $\Delta F$ at equilibrium. Away from
equilibrium---when one side has a concrete advantage---$\Delta F \neq PV$,
and the residual $\mathcal{E} = (\Delta F - PV)/\bar{T}$ measures the
advantage in piece units.

\section{Physical Interpretation}

The decomposition $G = U + PV - TS$ separates the evaluation into three
physically distinct channels:

\begin{enumerate}[nosep]
\item \textbf{Internal energy $U$} (x-ray mobility): The raw potential of
  each piece, determined by type and position on the board. Directly tracks
  material advantage. This is the dominant term.

\item \textbf{Entropic correction $TS$}: Subtracts the portion of the
  internal energy that is ``redundant''---overlapping x-ray coverage among
  friendly pieces, weighted by how concentrated the legal moves are.
  Penalizes piece clusters that cover the same squares. Small in magnitude
  relative to $U$ but captures important configurational effects.

\item \textbf{Potential energy $PV$}: Structural and tactical interactions
  learned from data. Captures coordination, king safety, pawn structure,
  spatial tension, and other features that $U - TS$ alone cannot express.
  This is where the engine learns the ``energetics'' of piece
  interactions, as distinct from the ``geometry'' captured by entropy.
\end{enumerate}

The distinction between $S$ and $PV$ mirrors a fundamental distinction in
statistical mechanics: \emph{entropic} interactions arise from phase-space
geometry (excluded volume, overlap of accessible states) and are captured
by $S$, while \emph{energetic} interactions arise from forces between
particles and are captured by $PV$. Both involve piece-piece relationships,
but through different thermodynamic channels.

\section{Summary}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Quantity} & \textbf{Definition} & \textbf{Chess meaning} \\
\midrule
$U$ & $\sum_i f_i$ (x-ray mobility) & Intrinsic piece potential \\
$T$ & $M / e^H$ (move concentration) & Game phase (cold $=$ middlegame) \\
$S$ & $1 - |\text{union}|/\sum f_i$ & X-ray redundancy among allies \\
$F$ & $U - TS$ & Usable energy \\
$PV$ & Fitted structural weights & Piece coordination and tactics \\
$\mathcal{E}$ & $(\Delta F - PV)/\bar{T}$ & Evaluation in piece units \\
\bottomrule
\end{tabular}
\end{center}

The framework provides a principled decomposition of chess evaluation into
interpretable thermodynamic components, with a clear fitting procedure
grounded in the equilibrium assumption. The potential energy weights are
the only learned parameters; all other quantities ($U$, $T$, $S$, $F$)
are computed analytically from the position.

\end{document}
